Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-05-19 main.Rmd
harmony clash notation changed

\noindent All cyclic granularities can be expressed in terms of the index set, and hence, we consider the data structure in Figure \@ref(fig:data-structure) for exploration of temporal data of this kind. This is a two-dimensional array extending the columns of tsibble by including the cyclic granularities $C_i$. The total number of cyclic granularities would be based on the number of linear granularities considered in the hierarchy table and presence of any aperiodic cyclic granularities. Suppose we have $n$ periodic linear granularities in the hierarchy table, $n(n-1)/2$ circular or quasi-circular cyclic granularities could be constructed. Let $N_C$ be the total number of contextual circular, quasi-circular and aperiodic granularities that could originate from $N_L$ periodic and aperiodic linear granularities. Any attempt to encode all or many of these cyclic granularities at the same time to develop insights on periodicity might fail or otherwise become too numerous for comprehensive human consumption. Instead, this big problem is broken down by focusing on pair of cyclic granularities ($C_i$, $C_j$) at a time and data sets of the form <$C_i$, $C_j$, $v$> forms the basis for exploration and analysis of the measured variable $v$.

@@

The way cyclic granularities relate become important when we consider the data structure in Figure \@ref(fig:data-structure). Let us consider two cyclic granularities $C_i$ and $C_j$, such that $C_i$ maps index set to a set $\{A_1, A_2, A_3, \dots, A_I$\}, and $C_j$ maps index set to a set $\{B_1, B_2, B_3, \dots, B_J$\}. Here, $A_k$'s or $B_l$'s are the levels/categories corresponding to $C_i$ and $C_j$ respectively. Let $S_{kl}$ be a subset of the index set such that for all $s \in S_{kl}$, $C_i(s) = A_k$ and $C_j(s) = B_l$. Data subsets for each combination of levels ($A_k$, $B_l$) like <$A_k$, $B_l$, $v(s)$> can be obtained for all $i \in {1, 2, \dots, l}$ and $j \in {1, 2, \dots, m}$  which will lead to $lm$ data subsets. Now, some situations can lead to few or many of these sets being empty. We will discuss few cases, where one or more of these $lm$ sets will be empty either due to the structure of the calendar, duration and location of events in a calendar or just by the construction of the cyclic granularities. <!--The implication of these empty sets on visualization will be discussed in Section \ref{sec:vis-interaction}. -->

@@

Firstly, empty combinations can arise due to the structure of the calendar or hierarchy. These are called "structurally" empty combinations. Let us take a specific example, where $C_i$ be day-of-month with 31 levels and $C_j$ be week-of-month with 5 levels. There will be $31\times 5=155$ sets $S_{kl}$ corresponding to possible combinations of $C_i$ and $C_j$. Many of these like $S_{1,5}$, $S_{21,2}$ are empty. This is also intuitive since the first day of the month can never correspond to fifth week of the month. Hence the pair (day-of-month, week-of-month) is a clash.

@@

Secondly, empty combinations can turn up due to differences in event location or duration in a calendar. These are called "event-driven" empty combinations. Let us consider $C_i$ be day-of-week with 7 levels and $C_j$ be WorkingDay/NonWorkingDay with 2 levels. While potentially all of these 14 sets $S_{kl}$ can be non-empty (it is possible to have a public holiday on any day-of-week), in practice many of these will probably have very few observations. For example, there are few (if any) public holidays on Wednesdays or Thursdays in any given year in Melbourne, Australia.

@@

Thirdly, empty combinations can be a result of how granularities are constructed. These are called "build-based" empty combinations. Let $C_i$ be Business-days, which are days from Monday to Friday except holidays and $C_j$ be day-of-month. Then the days denoting weekends in a month would not correspond to any Business days. This is different from structurally empty combinations because structure of the calendar does not lead to these missing combinations, but the construction of the granularities does.

An example when there will be no empty combinations could be where $C_i$ and $C_j$ maps index set to day-of-week and month-of-year respectively. Here $C_i$ can have 7 levels while $C_j$ can have 12 levels. So there are $12\times7=84$ sets $S_{kl}$. All of these are non-empty because every day-of-week can occur in every month. Hence, the pair (day-of-week, month-of-year) is a harmony.

>>>>

\noindent All cyclic granularities can be expressed in terms of the index set, and hence, we consider the data structure in Figure \@ref(fig:data-structure) for exploration of temporal data of this kind. This is a two-dimensional array extending the columns of tsibble by including the cyclic granularities. The total number of cyclic granularities would be based on the number of linear granularities considered in the hierarchy table and presence of any aperiodic cyclic granularities, for example, if we have $n$ periodic linear granularities in the hierarchy table, $n(n-1)/2$ circular or quasi-circular cyclic granularities could be constructed. Let $N_C$ be the total number of contextual circular, quasi-circular and aperiodic cyclic granularities that could originate from $N_L$ periodic and aperiodic linear granularities. Any attempt to encode all or many of these cyclic granularities at the same time to develop insights on periodicity might fail or otherwise become too numerous for comprehensive human consumption. Instead, this big problem is broken down by focusing on pair of cyclic granularities ($C_i$, $C_j$) at a time for $i, j \in \{1, 2, \dots, N_C\}$. Data sets of the form <$C_i$, $C_j$, $v$> then forms the basis for exploration and analysis of the measured variable $v$.

@@

The way cyclic granularities relate become important when we consider the data structure in Figure \@ref(fig:data-structure). Let us consider two cyclic granularities $C_i$ and $C_j$, such that $C_i$ maps index set to a set $\{A_k | k \in \mathbb{N}, k \leq K\}$ and $C_j$ maps index set to a set $\{B_l | l \in \mathbb{N}, l \leq L\}$. Here, $A_k$'s or $B_l$'s are the levels/categories corresponding to $C_i$ and $C_j$ respectively. Let $S_{kl}$ be a subset of the index set such that for all $s \in S_{kl}$, $C_i(s) = A_k$ and $C_j(s) = B_l$. Data subsets for each combination of levels ($A_k$, $B_l$) like <$A_k$, $B_l$, $v(s)$> can be obtained for all $k \in {1, 2, \dots, K}$ and $l \in {1, 2, \dots, L}$  which will lead to $KL$ data subsets. Now, some situations can lead to few or many of these sets being empty. We will discuss few cases, where one or more of these $KL$ sets will be empty either due to the structure of the calendar, duration and location of events in a calendar or just by the construction of the cyclic granularities. <!--The implication of these empty sets on visualization will be discussed in Section \ref{sec:vis-interaction}. -->

@@

\noindent Firstly, empty combinations can arise due to the structure of the calendar or hierarchy. These are called "structurally" empty combinations. Let us take a specific example, where $C_i$ be day-of-month with 31 levels and $C_j$ be week-of-month with 5 levels. There will be $31\times 5=155$ sets $S_{kl}$ corresponding to possible combinations of $C_i$ and $C_j$. Many of these like $S_{1,5}$, $S_{21,2}$ are empty. This is also intuitive since the first day of the month can never correspond to fifth week of the month. Hence the pair (day-of-month, week-of-month) is a clash.

@@

\noindent Secondly, empty combinations can turn up due to differences in event location or duration in a calendar. These are called "event-driven" empty combinations. Let us consider $C_i$ be day-of-week with 7 levels and $C_j$ be WorkingDay/NonWorkingDay with 2 levels. While potentially all of these 14 sets $S_{kl}$ can be non-empty (it is possible to have a public holiday on any day-of-week), in practice many of these will probably have very few observations. For example, there are few (if any) public holidays on Wednesdays or Thursdays in any given year in Melbourne, Australia.

@@

\noindent Thirdly, empty combinations can be a result of how granularities are constructed. These are called "build-based" empty combinations. Let $C_i$ be Business-days, which are days from Monday to Friday except holidays and $C_j$ be day-of-month. Then the days denoting weekends in a month would not correspond to any Business days. This is different from structurally empty combinations because structure of the calendar does not lead to these missing combinations, but the construction of the granularities does.

\noindent An example when there will be no empty combinations could be where $C_i$ and $C_j$ maps index set to day-of-week and month-of-year respectively. Here $C_i$ can have 7 levels while $C_j$ can have 12 levels. So there are $12\times7=84$ sets $S_{kl}$. All of these are non-empty because every day-of-week can occur in every month. Hence, the pair (day-of-week, month-of-year) is a harmony.
